Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-08-03 submission/manuscript.Rmd
update intro with Ada suggestions

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment has been attributed to being a larger risk factor for CRC [@Lichtenstein2000]. These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011]. Many of these environmental risk factors, including the microbiota, could be easily modifiable, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018]. Many of these previous case/control studies have identified resident bacterial taxa to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018]. Many of the bacterial species from these resident taxa actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Ultimately, these SCFAs could be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an anti-tumorigenic agent. Specific SCFAs, like butyrate, have shown positive results within model systems [@test_OKeefe2016]. For example, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018]. Additionally, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018]. These exciting results in model systems suggests that supplementation with food sources that bacteria use to create these SCFAs may be able to confer beneficial effects against CRC. 
However, it is important to note that these model systems provide only preliminary evidence towards the ability of SCFAs to reduce and treat tumors and the studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it tests previous case/control findings on SCFA concentrations in individuals with and without tumors. It also tests previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity worsens by increasing the number of patients with adenomas within the study. Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma). We also investigate whether OTUs that are most important to these models are closely associated with the classification of SCFA concentrations. Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. We also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate if there were any differences in genes involved with SCFA synthesis between individuals without tumors, patients with adenomas, and patients with carcinomas. Next, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without tumors and patients with adenomas or carcinomas. Additionally, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002]. We also analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations. Collectively, this investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US and the prevelance is increasing in invidiauls under 50 years of age [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment is considered a larger risk factor for CRC [@Lichtenstein2000]. These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011]. Many of these environmental risk factors, including the microbiota, are modifiable. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018]. Many of these previous case/control studies have identified resident bacterial taxa to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018]. Many of the bacterial species from these resident taxa identified in these case/control studies actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017]. The most extensively studied SCFAs are acetate, butyrate, and propionate [@Louis2016]. These SCFAs are hypothesized to be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an anti-tumorigenic agent. Specific SCFAs have shown positive results within model systems [@test_OKeefe2016]. For example, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018]. Additionally, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018]. These exciting results in model systems suggest that supplementation with food sources that bacteria use to create these SCFAs may confer beneficial effects against CRC. However, it is important to note that these model systems provide only preliminary evidence towards the ability of SCFAs to reduce and treat tumors and the studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, by using a seperate cohort it tests previous case/control findings on SCFA concentrations in individuals with and without tumors. Additionally, prior investigations grouped patients with adenoma and individuals without tumors into a single group to compare against patients with carcinomas. Despite doing this, the suggestion was made that a reduction in SCFA concentration would also be observed between patients with adenomas and indivdiuals without tumors. We also can test this suggestion because we have a larger number of patients with adenomas within our study and do not need to group patients with adenomas and individuals without tumors into a single group. Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma). We also investigate whether OTUs that are most important to these models are closely associated with the classification of SCFA concentrations. Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. To provide further support for our SCFA findings we also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate differences in genes involved with SCFA synthesis between individuals without tumors, patients with adenomas, and patients with carcinomas. Next, we investigated whether taxa associated with SCFA production were important to disease classification models. First, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without tumors and patients with adenomas or carcinomas. Second, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002]. Third, we analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations. Finally, we compared whether SCFA concentrations replace important taxa in disease models and whether the same OTUs are the most important variables used to classify disease and SCFA concentration. Collectively, this investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.

### 2018-08-03 submission/manuscript.tex
update intro with Ada suggestions

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the environment has been attributed to being a
larger risk factor for CRC (3). These environmental risk factors include
but are not limited to smoking cigarettes, diet, and the microbiota
(4--6). Many of these environmental risk factors, including the
microbiota, could be easily modifiable, and this has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(7--9) and whether the bacterial community is altered in CRC (10, 11).
Many of these previous case/control studies have identified resident
bacterial taxa to be decreased in patients with carcinoma tumors
(11--13). Many of the bacterial species from these resident taxa
actively produce short-chain fatty acids (SCFAs) from fiber that are
part of our general diet (14). The most extensively studied of these
SCFAs are acetate, butyrate, and propionate (15). Ultimately, these
SCFAs could be the main metabolites involved with protection against
tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an
anti-tumorigenic agent. Specific SCFAs, like butyrate, have shown
positive results within model systems (16). For example, butyrate has
been shown to inhibit cancer cell growth in \emph{in vitro} systems
(17). Additionally, fiber supplementation in mouse models of CRC caused
an overall reduction in tumor burden while also increasing SCFA
concentrations (18). These exciting results in model systems suggests
that supplementation with food sources that bacteria use to create these
SCFAs may be able to confer beneficial effects against CRC. However, it
is important to note that these model systems provide only preliminary
evidence towards the ability of SCFAs to reduce and treat tumors and the
studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas or individuals
without colon tumors (19). Although this would argue that increasing
SCFA concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence (20). These findings would argue
against the utility of treatments that aim to use SCFAs to reduce or
protect against tumorigenesis. Given the lack of clear evidence in human
studies of the benefit of SCFAs in CRC, there is a need for more
investigation into this area.

Our study fills some of the current gaps in the literature that relate
to the study of SCFAs and CRC in human populations. Specifically, it
tests previous case/control findings on SCFA concentrations in
individuals with and without tumors. It also tests previous suggestions
that there is a continuous reduction in SCFA concentrations as tumor
severity worsens by increasing the number of patients with adenomas
within the study. Additionally, we build upon these observations by
assessing the utility of using SCFAs and Operational Taxonomic Units
(OTUs) as a risk stratification tool of colorectal tumors (adenoma or
carcinoma). We also investigate whether OTUs that are most important to
these models are closely associated with the classification of SCFA
concentrations. Collectively, this study provides important information
on the replicability of previous findings in humans by extensively
studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors. We also
used PICRUSt (21) and metagenomic sequencing to investigate if there
were any differences in genes involved with SCFA synthesis between
individuals without tumors, patients with adenomas, and patients with
carcinomas. Next, using the cross-sectional data, we analyzed the number
of correlations between OTU relative abundance and SCFA concentrations
across individuals without tumors and patients with adenomas or
carcinomas. Additionally, we assessed the affect adding SCFA
concentrations to OTU data had on classification of patients with
adenomas or carcinomas using the Random Forest algorithm (22). We also
analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
Collectively, this investigation provides additional information as to
whether SCFAs are decreased in patients with colorectal tumors and
provides context as to whether targeting taxa to increase SCFA
concentrations is a viable option to protect against colon
tumorigenesis.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US and the prevelance is increasing in
invidiauls under 50 years of age (1, 2). Although there is a genetic
component to the disease, the environment is considered a larger risk
factor for CRC (3). These environmental risk factors include but are not
limited to smoking cigarettes, diet, and the microbiota (4--6). Many of
these environmental risk factors, including the microbiota, are
modifiable. This has lead to the investigation of how the microbiota may
exacerbate or cause tumorigensis (7--9) and whether the bacterial
community is altered in CRC (10, 11). Many of these previous
case/control studies have identified resident bacterial taxa to be
decreased in patients with carcinoma tumors (11--13). Many of the
bacterial species from these resident taxa identified in these
case/control studies actively produce short-chain fatty acids (SCFAs)
from fiber that are part of our general diet (14). The most extensively
studied SCFAs are acetate, butyrate, and propionate (15). These SCFAs
are hypothesized to be the main metabolites involved with protection
against tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an
anti-tumorigenic agent. Specific SCFAs have shown positive results
within model systems (16). For example, butyrate has been shown to
inhibit cancer cell growth in \emph{in vitro} systems (17).
Additionally, fiber supplementation in mouse models of CRC caused an
overall reduction in tumor burden while also increasing SCFA
concentrations (18). These exciting results in model systems suggest
that supplementation with food sources that bacteria use to create these
SCFAs may confer beneficial effects against CRC. However, it is
important to note that these model systems provide only preliminary
evidence towards the ability of SCFAs to reduce and treat tumors and the
studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas or individuals
without colon tumors (19). Although this would argue that increasing
SCFA concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence (20). These findings argue against
the utility of treatments that aim to use SCFAs to reduce or protect
against tumorigenesis. Given the lack of clear evidence in human studies
of the benefit of SCFAs in CRC, there is a need for more investigation
into this area.

Our study fills some of the current gaps in the literature that relate
to the study of SCFAs and CRC in human populations. Specifically, by
using a seperate cohort it tests previous case/control findings on SCFA
concentrations in individuals with and without tumors. Additionally,
prior investigations grouped patients with adenoma and individuals
without tumors into a single group to compare against patients with
carcinomas. Despite doing this, the suggestion was made that a reduction
in SCFA concentration would also be observed between patients with
adenomas and indivdiuals without tumors. We also can test this
suggestion because we have a larger number of patients with adenomas
within our study and do not need to group patients with adenomas and
individuals without tumors into a single group. Additionally, we build
upon these observations by assessing the utility of using SCFAs and
Operational Taxonomic Units (OTUs) as a risk stratification tool of
colorectal tumors (adenoma or carcinoma). We also investigate whether
OTUs that are most important to these models are closely associated with
the classification of SCFA concentrations. Collectively, this study
provides important information on the replicability of previous findings
in humans by extensively studying how SCFAs are associated with
colorectal tumors.

To accomplish this task we directly measured the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors. To
provide further support for our SCFA findings we also used PICRUSt (21)
and metagenomic sequencing to investigate differences in genes involved
with SCFA synthesis between individuals without tumors, patients with
adenomas, and patients with carcinomas. Next, we investigated whether
taxa associated with SCFA production were important to disease
classification models. First, using the cross-sectional data, we
analyzed the number of correlations between OTU relative abundance and
SCFA concentrations across individuals without tumors and patients with
adenomas or carcinomas. Second, we assessed the affect adding SCFA
concentrations to OTU data had on classification of patients with
adenomas or carcinomas using the Random Forest algorithm (22). Third, we
analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
Finally, we compared whether SCFA concentrations replace important taxa
in disease models and whether the same OTUs are the most important
variables used to classify disease and SCFA concentration. Collectively,
this investigation provides additional information as to whether SCFAs
are decreased in patients with colorectal tumors and provides context as
to whether targeting taxa to increase SCFA concentrations is a viable
option to protect against colon tumorigenesis.
